%%% This needs to be actually written in Latex. Just throwing in some comments for now.


Issues with Python:

So one of the road blocks while getting these sorts implemented and our tests run, and our
data gathered, was with Python's limitations. Mainly I speak of the way Python handles
recursion. The very first hurdle discovered was that Python has a recursion depth limit
of 1000. This "prevents infinite recursion from causing an overflow of the C stack and 
crashing Python." (http://docs.python.org/library/sys.html)

This hurdle can be overcome by setting the limit manually, but as the warning states
Python can crash. By setting the number to one million, I was able to run the Python sort
which was doing the recursion. However, the next road block was a Segmentation Fault.
This lead me to believe that the implementation was not correct for Python. I began to
talk to peers and search the internet, and found a site where someone had implemented
a tail-recursion function. Apparently, Python does not normally do tail recursion.
It was difficult to follow the example on the site, though. The algorithm I was
implementing was concatenating a list with the return value of the recursive call,
and Python thought I wanted to concatenate a list with a function.

After spending more time speaking with peers on the subject and researching more,
I decided to pull the plug on Python, for doing recursive sorting algorithms. I am
researching things people have done to work around this, including Stackless Python
and C Python. A very interesting point, is that Stackless Python acknowledges on
their site that when you implement these workarounds you are implementing a Python
paradigm shift. To change the paradigm of the language within a program is rather
significant. However, for this project and the time frame we have been given, the
goal is not to turn this into a Golden Hammer. Python is not the correct tool for
doing a recursive sort of a list, so the focus will instead be on the correct tools.

Appendix: Stackless Python
Stackless Python is an implementation of Python that decouples the C stack and thus
skirts the recursion issues of Python. There is an excellent paper posted on their
website here: http://www.stackless.com/spcpaper.htm
